---
title: "Python and R"
description: >-
  Setting up Python and R properly — interpreters, package managers, and the environment to create before writing a line of analysis.
---

## Setting Up Python

Python is the language most of the infrastructure in this book is written in and expects. NumPy, pandas and scikit-learn cover the numerical and modelling work; almost everything else — web frameworks, cloud SDKs, orchestration tools — has a Python client whether or not it has an R one. That reach is the practical argument for it, more than anything about the language itself.

### Installing Python

You have a few reasonable ways to install Python for data science. Each has trade-offs:

- **Miniforge** (recommended for most readers): a minimal, community-maintained conda distribution that defaults to `conda-forge`, a free package repository — conda calls these "channels" — maintained by the community rather than by a company. Gives you `conda` for environment management without the licensing concerns that now apply to the full Anaconda distribution for commercial users in larger organisations.
- **Anaconda**: the "batteries-included" distribution. Convenient, but its Terms of Service have restricted free commercial use for organisations with 200 or more employees since 2020, and a 2024 update extended that to government bodies and non-profits of the same size. Fine for learning and personal projects.
- **`uv`** (the modern, fast option): a single binary that installs Python versions and manages virtual environments and packages. Much faster than conda or pip and rapidly becoming the default for pure-Python projects. A great choice once you're comfortable, especially if you don't need conda's scientific package ecosystem.
- **System Python** (`python.org` installer, `brew install python`, or `apt install python3`): simple, but you'll still want to create isolated environments per project with `venv` or `uv venv`.

For this book we'll use conda-based instructions since they work uniformly across Windows, macOS, and Linux and handle tricky scientific packages well. Everything shown below works with either Miniforge or Anaconda.

#### Installing Miniforge (or Anaconda)

1. Visit the [Miniforge releases page](https://github.com/conda-forge/miniforge/releases/latest) (or the [Anaconda download page](https://www.anaconda.com/download) if you prefer)
2. Download the appropriate installer for your operating system
3. Run the installer and follow the prompts

During installation on Windows, you may be asked whether to add the installation to your PATH environment variable. Leave this unchecked and use the Miniforge Prompt (or Anaconda Prompt) from the Start menu, which avoids conflicts with other Python installations on your system.

The "PATH" is like an address book that tells your computer where to find programs when you type their names. Adding your conda install to PATH means you can use Python from any command prompt, but it could cause conflicts with other versions of Python on your system.

#### Verifying Installation

Open a new terminal (or Miniforge/Anaconda Prompt on Windows) and type:

```bash
python --version
```

You should see the Python version number. Also, check that conda is installed:

```bash
conda --version
```

:::{.callout-important}
## Don't install packages into the base installation

The next thing most people do is `pip install pandas` and carry on. Resist it. Packages installed globally accumulate, conflict, and eventually break each other in ways that are genuinely painful to unpick — and nothing records what you installed, so the analysis you write today may not run tomorrow.

Create a per-project environment first. That is the whole subject of the [Reproducible Environments](environments.qmd) chapter, and it's worth reading before you install your second package. The short version, using the conda you have just installed:

```bash
conda create -n myproject python=3.13
conda activate myproject
```

Install into that, not into `base`.
:::

### Using Jupyter Notebooks

A Jupyter notebook interleaves code, its output, and your prose in one document — a lab notebook you can re-run. It's where most exploratory Python work happens.

Miniforge is deliberately minimal, so Jupyter is not installed yet. Create an environment for working through this book and install what the next few chapters need:

```bash
# Create an environment named 'datasci' with the packages used below
conda create -n datasci python=3.13 numpy pandas matplotlib jupyter

# Activate it. Do this in every new terminal session where you want these packages.
conda activate datasci

# Launch Jupyter Notebook
jupyter notebook
```

The first command takes a few minutes — conda is resolving a consistent set of versions for everything you asked for and everything those packages depend on in turn. Later chapters refer back to this `datasci` environment; [Reproducible Environments](environments.qmd) explains why one environment per project beats one environment for everything.

This opens a web browser where you can create and work with notebooks. Let's create a simple notebook to verify everything works:

1. Click "New" → "Python 3"
2. In the first cell, type:

```{python}
#| eval: false
import numpy as np
import pandas as pd
import matplotlib.pyplot as plt

# Create some sample data
data = pd.DataFrame({
    'x': range(1, 11),
    'y': np.random.randn(10)
})

# Create a simple plot
plt.figure(figsize=(8, 4))
plt.plot(data['x'], data['y'], marker='o')
plt.title('Sample Plot')
plt.xlabel('X-axis')
plt.ylabel('Y-axis')
plt.grid(True)
plt.show()

print("Python environment is working correctly!")
```

3. Press Shift+Enter to run the cell

If you see a plot and the success message, your Python setup is working.

### Installing Additional Packages

You'll need more packages as you go. Two ways to install them:

```bash
# Using conda (preferred when available)
conda install package_name

# Using pip (when packages aren't available in conda)
pip install package_name
```

Prefer conda where the package has compiled C or C++ underneath it — which covers most of the numerical stack. Conda ships prebuilt binaries and resolves the system libraries they depend on; pip is more likely to try compiling on your machine and fail for want of a toolchain.

## Setting Up R

R was built for statistical computing rather than adapted to it, and it shows in small places: model objects that print something useful without being asked, missing values handled as a first-class idea, and CRAN packages written by the statisticians who developed the methods in them. The "Python or R?" section at the end of this chapter goes into when to reach for which.

### Installing R

Let's install both R itself and RStudio, a popular integrated development environment for R.

#### Installing Base R

1. Visit the [Comprehensive R Archive Network (CRAN)](https://cran.r-project.org/)
2. Click on the link for your operating system
3. Follow the installation instructions

#### Installing RStudio Desktop

RStudio provides a user-friendly interface for working with R.

1. Visit the [Posit RStudio download page](https://posit.co/download/rstudio-desktop/) (RStudio's parent company rebranded to Posit in 2022)
2. Download the free RStudio Desktop version for your operating system
3. Run the installer and follow the prompts

Think of R as the engine and RStudio as the dashboard that makes it easier to control that engine. You could use R without RStudio, but RStudio makes many tasks more convenient.

#### Verifying Installation

Open RStudio and enter this command in the console (lower-left pane):

```{r}
#| eval: false
R.version.string
```

RStudio also prints the same information unprompted at the start of every new session, so you may already see it above your cursor. It should look something like this:

`R version 4.6.1 (2026-06-24 ucrt) -- "Happy Hop"`
`Copyright (C) 2026 The R Foundation for Statistical Computing`
`Platform: x86_64-w64-mingw32/x64`

Your version number and codename will differ — R releases a new minor version each April and patches through the year. The `ucrt` and `Platform` lines only appear on Windows.

### The R Packages to Start With

Five packages cover most of what the rest of this book does in R:

```{r}
#| eval: false
install.packages(c("tidyverse", "rmarkdown", "shiny", "knitr", "plotly"))
```

This installs:

- **tidyverse**: A collection of packages for data manipulation and visualisation
- **rmarkdown**: For creating documents that mix code and text
- **shiny**: For building interactive web applications
- **knitr**: For dynamic report generation
- **plotly**: For interactive visualisations

The tidyverse is the one to learn first. It replaces most of base R's data manipulation with a consistent set of verbs, and the other four assume you have it.

### Creating Your First R Script

Verify the setup with a short script:

1. In RStudio, go to File → New File → R Script
2. Enter the following code:

```{r}
#| eval: false
# Load libraries
library(tidyverse)

# Create sample data
data <- tibble(
  x = 1:10,
  y = rnorm(10)
)

# Create a plot with ggplot2
ggplot(data, aes(x = x, y = y)) +
  geom_point() +
  geom_line() +
  labs(title = "Sample Plot in R",
       x = "X-axis",
       y = "Y-axis") +
  theme_minimal()

print("R environment is working correctly!")
```

3. Click the "Run" button or press Ctrl+Enter (Cmd+Enter on Mac) to execute the code

If you see a plot in the lower-right pane and the success message in the console, your R setup is working.

### Understanding R Packages

Unlike Python, where conda or pip manage packages, R has its own built-in package management system accessed through functions like `install.packages()` and `library()`.

There are thousands of R packages available on CRAN, with more on Bioconductor (for bioinformatics) and GitHub. To install a package from GitHub, you first need the devtools package:

```{r}
#| eval: false
install.packages("devtools")
devtools::install_github("username/package")
```

Think of CRAN as the official app store for R packages, while GitHub is like getting apps directly from developers. Both are useful, but packages on CRAN have gone through more quality checks.

## Python or R?

The question comes up constantly and has a boring answer: both are good, the differences that once mattered mostly don't any more, and the deciding factor is almost always what the people around you use.

Where genuine differences remain:

- **R** is stronger for statistical modelling and for publication-quality graphics. The tidyverse is a more coherent data-manipulation vocabulary than pandas, and ggplot2 has no true equal in Python. If your work is econometrics, biostatistics, or survey analysis, R's package ecosystem is deeper.
- **Python** is stronger everywhere the analysis has to connect to something else — a web application, a production pipeline, a machine learning framework, a cloud service. If your output is a system rather than a paper, Python is the shorter path.

Both are covered throughout this book because most working analysts eventually use both. Quarto renders either. Git doesn't care. The environment and deployment chapters apply to both with minor syntax differences.

If you have no constraint and no preference: learn Python first, because more of the infrastructure in this book is Python-native, then pick up R when you meet a statistical problem it's better at.

## Conclusion

**By the end of this chapter you should have:**

- Python installed via Miniforge, with `python --version` and `conda --version` both working
- A conda environment named `datasci` containing numpy, pandas, matplotlib, and Jupyter, which later chapters assume
- R and RStudio installed, with `R.version.string` reporting a current version
- A Jupyter notebook that runs a cell and produces a plot
- An R script that loads ggplot2 and produces a plot
- Read the [Reproducible Environments](environments.qmd) chapter before installing much else

The next chapter covers the editor you'll work in and the version control that everything after it assumes.
