---
title: "Data Visualisation"
description: >-
  Building charts that communicate — the grammar of graphics, ggplot2 and matplotlib, and the choices that decide whether a plot lands.
---

## Why Visualisation Matters in Data Science

A chart works because it hands the reader a shape to recognise instead of a column of numbers to compare one pair at a time. That applies as much when the reader is you — most outliers and most data-entry mistakes are found by plotting the column, not by summarising it — as when it's a stakeholder who will act on the result.

The reporting chapter covered the *document* that wraps your analysis. This chapter covers the charts and diagrams that go inside it, in three categories:

- **Python visualisation libraries**: matplotlib, seaborn, plotly
- **R visualisation libraries**: ggplot2 and interactive wrappers
- **Code-based diagramming** with Mermaid for flowcharts, ER diagrams, and architecture sketches

Interactive web dashboards built on top of these libraries (Shiny, Dash, Streamlit) are covered in the [Web Development for Data Scientists](web_dev.qmd) chapter, since they're fundamentally running web applications rather than visualisations embedded in a document.

## Python Visualisation Libraries

Python has several established visualisation libraries, each with different strengths and use cases.

### Matplotlib: The Foundation

Matplotlib is the original Python visualisation library and serves as the foundation for many others. It provides precise control over every element of a plot.

```{python}
#| eval: false
import matplotlib.pyplot as plt
import numpy as np

# Generate data
x = np.linspace(0, 10, 100)
y = np.sin(x)

# Create a figure and axis
fig, ax = plt.subplots(figsize=(10, 6))

# Plot data
ax.plot(x, y, 'b-', linewidth=2, label='sin(x)')

# Add labels and title
ax.set_xlabel('X-axis', fontsize=14)
ax.set_ylabel('Y-axis', fontsize=14)
ax.set_title('Sine Wave', fontsize=16)

# Add grid and legend
ax.grid(True, linestyle='--', alpha=0.7)
ax.legend(fontsize=12)

# Save and show the figure
plt.savefig('sine_wave.png', dpi=300, bbox_inches='tight')
plt.show()
```

Matplotlib hands you a blank canvas and expects you to place every element yourself. That's more code than the alternatives for an ordinary chart, and the right trade when you need exact control over every line, tick and label — a figure for journal submission, say, where the specification is fixed.

### Seaborn: Statistical Visualisation

Seaborn builds on Matplotlib to provide high-level functions for common statistical visualisations.

```{python}
#| eval: false
import seaborn as sns
import pandas as pd
import matplotlib.pyplot as plt

# Set the theme
sns.set_theme(style="whitegrid")

# Load example data
tips = sns.load_dataset("tips")

# Create a visualisation
plt.figure(figsize=(12, 6))
sns.boxplot(x="day", y="total_bill", hue="smoker", data=tips, palette="Set3")
plt.title("Total Bill by Day and Smoker Status", fontsize=16)
plt.xlabel("Day", fontsize=14)
plt.ylabel("Total Bill ($)", fontsize=14)
plt.tight_layout()
plt.show()
```

Reach for Seaborn when the chart you want is a standard statistical view of a pandas DataFrame — box plot, violin plot, regression fit. It's usually one function call where the matplotlib equivalent is a dozen lines, and its defaults look presentable without adjustment.

### Plotly: Interactive Visualisations

Plotly creates interactive visualisations that can be embedded in web applications or Jupyter notebooks.

```{python}
#| eval: false
import plotly.express as px
import pandas as pd

# Load example data
df = px.data.gapminder().query("year == 2007")

# Create an interactive scatter plot
fig = px.scatter(
    df, x="gdpPercap", y="lifeExp", size="pop", color="continent",
    log_x=True, size_max=60,
    title="GDP per Capita vs Life Expectancy (2007)",
    labels={"gdpPercap": "GDP per Capita", "lifeExp": "Life Expectancy (years)"}
)

# Update layout
fig.update_layout(
    width=900, height=600,
    legend_title="Continent",
    font=dict(family="Arial", size=14)
)

# Show the figure
fig.show()
```

Zoom, pan and hover-for-detail come free, which matters when a reader wants to interrogate a chart rather than just look at it. Plotly exists in both Python and R with much the same API, so it's the one library here you don't have to learn twice.

### Choosing Between Python Libraries

A pragmatic decision tree for the non-CS reader:

- **Need absolute control or a publication-quality static figure?** Matplotlib.
- **Want a standard statistical plot from a DataFrame with one function call?** Seaborn.
- **Producing an HTML report or notebook where hover, zoom, and tooltips add value?** Plotly.

You'll usually end up with all three in your toolkit: seaborn for quick EDA, plotly for HTML-bound output, and matplotlib when one of the other two can't quite render what you want.

## R Visualisation Libraries

R's visualisation ecosystem is built around ggplot2, which is what the tidyverse, most textbooks, and most Stack Overflow answers assume.

### ggplot2: Grammar of Graphics

ggplot2 is the default choice for data visualisation in R, built on the Grammar of Graphics.

```{r}
#| eval: false
library(ggplot2)
library(dplyr)

# Load dataset
data(diamonds, package = "ggplot2")

# Create a sample of the data
set.seed(42)
diamonds_sample <- diamonds |>
  slice_sample(n = 1000)

# Create basic plot
p <- ggplot(diamonds_sample, aes(x = carat, y = price, color = cut)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Diamond Price vs. Carat by Cut Quality",
    subtitle = "Sample of 1,000 diamonds",
    x = "Carat (weight)",
    y = "Price (USD)",
    color = "Cut Quality"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12, color = "gray50"),
    axis.title = element_text(size = 12),
    legend.position = "bottom"
  )

# Display the plot
print(p)

# Save the plot
ggsave("diamond_price_carat.png", p, width = 10, height = 6, dpi = 300)
```

You build a ggplot2 chart in layers, joined with `+`. First the data and the aesthetic mappings — which variable becomes the x position, which becomes the colour — then geometric objects to draw them with, then any statistical transformation, scales, coordinate system and theme. Each layer is simple on its own, and a complicated chart is a longer stack of them rather than a different kind of code.

### Interactive R Visualisations

R also offers interactive visualisation libraries:

```{r}
#| eval: false
library(plotly)
library(dplyr)

# install.packages("gapminder") if you don't have it - it isn't part of
# the tidyverse, and nothing else in this book depends on it
data(gapminder, package = "gapminder")
data_2007 <- gapminder |>
  filter(year == 2007)

# Create interactive plot
p <- plot_ly(
  data = data_2007,
  x = ~gdpPercap,
  y = ~lifeExp,
  size = ~pop,
  color = ~continent,
  type = "scatter",
  mode = "markers",
  sizes = c(5, 70),
  marker = list(opacity = 0.7, sizemode = "diameter"),
  hoverinfo = "text",
  text = ~paste(
    "Country:", country, "<br>",
    "Population:", format(pop, big.mark = ","), "<br>",
    "Life Expectancy:", round(lifeExp, 1), "years<br>",
    "GDP per Capita:", format(round(gdpPercap), big.mark = ","), "USD"
  )
) |>
  layout(
    title = "GDP per Capita vs. Life Expectancy (2007)",
    xaxis = list(
      title = "GDP per Capita (USD)",
      type = "log",
      gridcolor = "#EEEEEE"
    ),
    yaxis = list(
      title = "Life Expectancy (years)",
      gridcolor = "#EEEEEE"
    ),
    legend = list(title = list(text = "Continent"))
  )

# Display the plot
p
```

The R version of plotly can convert ggplot2 visualisations to interactive versions with a single function call:

```{r}
#| eval: false
# Convert a ggplot to an interactive plotly visualisation
ggplotly(p)
```

Because `ggplotly()` takes a finished ggplot, you can build the chart once in syntax you already know and decide about interactivity afterwards. That's the escape hatch when the same analysis has to ship as both a PDF and a web page.

## Code-Based Diagramming with Mermaid

Workflows, architectures and schemas are much easier to show than to describe. Rather than drawing them in a separate tool and pasting in an image, you can write them as code in the Quarto document itself, using Mermaid.

### Why Use Mermaid for Data Science?

A diagram drawn in a drawing tool becomes a PNG somebody has to find, open, edit, re-export and re-insert. Six months later the original file is on a laptop that's been replaced and the diagram is quietly wrong. A Mermaid diagram is a few lines of text in the document itself: it diffs in Git, it renders when the document renders, and correcting it means editing a line.

### Creating Mermaid Diagrams in Quarto

Quarto has built-in support for Mermaid diagrams. To create a diagram, use a code block with the `mermaid` engine:

```{mermaid}
flowchart LR
    A[Raw Data] --> B[Data Cleaning]
    B --> C[Exploratory Analysis]
    C --> D[Feature Engineering]
    D --> E[Model Training]
    E --> F[Evaluation]
    F --> G[Deployment]
```

The syntax starts with the diagram type (`flowchart`), followed by the direction (`LR` for left-to-right), and then the definition of nodes and connections.

### Diagram Types for Data Science

Mermaid supports several diagram types that are particularly useful for data science:

#### Flowcharts

Flowcharts are perfect for documenting data pipelines and analysis workflows:

```{mermaid}
flowchart TD
    A[Raw Data] --> B{Missing Values?}
    B -->|Yes| C[Imputation]
    B -->|No| D[Feature Engineering]
    C --> D
    D --> E[Train Test Split]
    E --> F[Model Training]
    F --> G[Evaluation]
    G --> H{Performance<br>Acceptable?}
    H -->|Yes| I[Deploy Model]
    H -->|No| J[Tune Parameters]
    J --> F
```

This top-down (TD) flowchart illustrates a complete machine learning workflow with decision points. Notice how you can use different node shapes (rectangles, diamonds) and add text to connections.

#### Class Diagrams

Class diagrams help explain data structures and relationships:

```{mermaid}
classDiagram
    class Dataset {
        +DataFrame data
        +load_from_csv(filename)
        +split_train_test(test_size)
        +normalize()
    }

    class Model {
        +train(X, y)
        +predict(X)
        +evaluate(X, y)
        +save(filename)
    }

    class Pipeline {
        +steps
        +add_step(transformer)
        +fit_transform(data)
    }

    Dataset --> Model: provides data to
    Pipeline --> Dataset: processes
    Pipeline --> Model: feeds into
```

This diagram shows the relationships between key classes in a machine learning system. It's useful for documenting the architecture of your data science projects.

#### Sequence Diagrams

Sequence diagrams show interactions between components over time:

```{mermaid}
sequenceDiagram
    participant U as User
    participant API as REST API
    participant ML as ML Model
    participant DB as Database

    U->>API: Request prediction
    API->>DB: Fetch features
    DB-->>API: Return features
    API->>ML: Send features for prediction
    ML-->>API: Return prediction
    API->>DB: Log prediction
    API-->>U: Return results
```

This diagram illustrates the sequence of interactions in a model deployment scenario, showing how data flows between the user, API, model, and database.

#### Gantt Charts

Gantt charts are useful for project planning and timelines:

```{mermaid}
gantt
    title Data Science Project Timeline
    dateFormat YYYY-MM-DD

    section Data Preparation
    Collect raw data       :a1, 2026-01-01, 10d
    Clean and validate     :a2, after a1, 5d
    Exploratory analysis   :a3, after a2, 7d
    Feature engineering    :a4, after a3, 8d

    section Modeling
    Split train/test       :b1, after a4, 1d
    Train baseline models  :b2, after b1, 5d
    Hyperparameter tuning  :b3, after b2, 7d
    Model evaluation       :b4, after b3, 4d

    section Deployment
    Create API            :c1, after b4, 6d
    Documentation         :c2, after b4, 8d
    Testing               :c3, after c1, 5d
    Production release    :milestone, after c2 c3, 0d
```

This Gantt chart shows the timeline of a data science project, with tasks grouped into sections and dependencies between them clearly indicated.

#### Entity-Relationship Diagrams

ER diagrams are valuable for database schema design:

```{mermaid}
erDiagram
    CUSTOMER ||--o{ ORDER : places
    ORDER ||--|{ ORDER_ITEM : contains
    PRODUCT ||--o{ ORDER_ITEM : "ordered in"
    CUSTOMER {
        int customer_id PK
        string name
        string email
        date join_date
    }
    ORDER {
        int order_id PK
        int customer_id FK
        date order_date
        float total_amount
    }
    ORDER_ITEM {
        int order_id PK,FK
        int product_id PK,FK
        int quantity
        float price
    }
    PRODUCT {
        int product_id PK
        string name
        string category
        float unit_price
    }
```

This diagram shows a typical e-commerce database schema with relationships between tables and their attributes.

### Styling Mermaid Diagrams

You can customise the appearance of your diagrams:

::: {.content-visible when-format="html"}

```{mermaid}
flowchart LR
    A[Data Collection] --> B[Data Cleaning]
    B --> C[Analysis]

    style A fill:#f9f,stroke:#333,stroke-width:2px
    style B fill:#bbf,stroke:#33f,stroke-width:2px
    style C fill:#bfb,stroke:#3f3,stroke-width:2px
```

:::

This diagram uses custom colours and border styles for each node to highlight different stages of the process.

### Generating Diagrams Programmatically

For complex or dynamic diagrams, you can generate Mermaid code programmatically. The trick is `output: asis`, which tells Quarto to drop the chunk's output into the document as raw Markdown rather than as printed R output — so a Mermaid fence you build with `cat()` is processed exactly as if you'd typed it:

```{r}
#| label: dynamic-mermaid
#| output: asis

# Define the steps in a data pipeline
steps <- c("Import Data", "Clean Data", "Feature Engineering",
           "Split Dataset", "Train Model", "Evaluate", "Deploy")

# Generate Mermaid flowchart code
mermaid_code <- c(
  "```{mermaid}",
  "flowchart LR"
)

# Add connections between steps
for (i in 1:(length(steps)-1)) {
  mermaid_code <- c(
    mermaid_code,
    sprintf("    %s[\"%s\"] --> %s[\"%s\"]",
            LETTERS[i], steps[i],
            LETTERS[i+1], steps[i+1])
  )
}

mermaid_code <- c(mermaid_code, "```")

# Output the Mermaid code
cat(paste(mermaid_code, collapse = "\n"))
```

The diagram above was generated by that code when this book was rendered — change the `steps` vector and the flowchart changes with it. That's the point: a diagram derived from data or configuration can't drift out of sync with the thing it documents, which is the usual fate of a hand-drawn one.

### Best Practices for Diagrams in Data Science

A diagram should carry weight the prose would otherwise have to. The test is whether removing it would cost the reader anything — if the surrounding paragraphs already say the same thing, you have decoration.

The examples above each have a sentence underneath explaining what they show, which is a teaching device for readers meeting Mermaid syntax for the first time. In your own documents, don't. If a diagram needs a paragraph telling the reader how to read it, simplify the diagram.

Beyond that: keep one diagram to one idea, use the same shapes and direction across a project so readers aren't relearning your conventions each time, and edit the diagram in the same commit as the code it describes.

## Conclusion

The libraries covered here — matplotlib, seaborn, plotly, ggplot2 — are the workhorses for the static and lightly-interactive charts you embed in reports. Mermaid handles the architecture and workflow diagrams that explain how your system fits together rather than what the data looks like.

**By the end of this chapter you should have:**

- A plotting library installed and working in each language you use — matplotlib or plotly for Python, ggplot2 for R
- One chart saved to a file at a resolution suitable for a report, rather than only ever viewed on screen
- One interactive chart embedded in a rendered HTML document, with hover and zoom working
- A Mermaid diagram rendering inside a Quarto document, so you can draw a pipeline without opening a drawing tool

As your datasets and models grow, you'll eventually hit the limits of what your laptop can render or compute. The next chapter steps away from the document and visualisation layer to look at cloud platforms: where to run notebooks and training jobs when local hardware isn't enough, and how to do so without accidentally running up a bill. After that, we return to the "how do I put this in front of a stakeholder" question with web development and interactive dashboards.
